---
title: "Generate day/night transition mapping"
---

```{r}
devtools::load_all()
```

We need to evaluate the day/night status for each pixel in the raster. We will fix to UTC so that we can easily compare to the current time on the watch. The steps are:

- Generate a raster to lat/lon coordinate converter
- Generate a grid of times - middle of every month for the next 12 months
  - we presume the yearly variation is tiny
  - the bigger issue is the within-month variability but hey, we are also doing this at an extremely coarse resolution so *shrugs*
- Calculate the sunrise/sunset time for each pixel in the raster
- Convert this into the most minimal representation for the watch to recreate the day/night transition map

We will use `suntools::sunriset()` - takes coords, datetime, crs, direction (sunrise/sunset), and whether to return POSIXct.

```{r}
day_image_path <- here::here("blue_pixel","resources","blue_marble.png")
day_image <- magick::image_read(day_image_path)
width <- magick::image_info(day_image)$width
height <- magick::image_info(day_image)$height

# get the available pixels mask (TRUE = non-transparent)
raster_data <- day_image |> as.raster()
# check for transparent pixels via "transparent" string
available_pixels_mask <- raster_data != "transparent"

# for each row, get the min max column index of available pixels
min_max_coords <- apply(available_pixels_mask, 1, function(row) {
    available_indices <- which(row)
    if (length(available_indices) == 0) {
        return(c(NA, NA))
    } else {
        return(c(min(available_indices), max(available_indices)))
    }
}) |> t() |> as.data.frame() |> dplyr::rename(min_x = V1, max_x = V2)

# assert min_max_coords is of length height
stopifnot(nrow(min_max_coords)  == height)
```

TODO: I will need to ensure the conversion from circle to rect image keeps the earth as cirular - so find the min dimension, and resize to that and then pad to the height dim.

```{r}
# calculate the radius
radius <- min_max_coords |> dplyr::mutate(
    dist = (max_x - min_x)
) |> dplyr::select(dist) |> max(na.rm=T)

project_to_lat_lon(
    x = 157-width/2,
    y = height/2-168,
    lat_ref = -3.84,
    lon_ref = 133.96,
    radius = radius
)
```

```{r}
# show the image as a plot
library(ggplot2)
plot_data <- available_pixels_mask |> 
    as.data.frame() |> 
    dplyr::mutate(y = dplyr::row_number()) |>
    tidyr::pivot_longer(-y, names_to = "x", values_to = "value") |> 
    dplyr::mutate(x = as.integer(gsub("V", "", x)))

ggplot() +
    geom_raster(
        data = plot_data,
        aes(x = x, y = y, fill = value)
    ) +
    scale_fill_manual(values = c("FALSE" = "black", "TRUE" = "blue")) +
    theme_void() +
    theme(legend.position = "none") +
    coord_fixed()
```

```{r}
# Get one time per month for the next 12 months
datetime_per_month <- lubridate::ymd_hms("2026-01-01 12:00:00", tz = "UTC") + lubridate::dmonths(0.5) + lubridate::dmonths(0:11)

# Create a data.frame to store the sunrise/sunset times for each month
sunrise_sunset_df <- data.frame(
    datetime = datetime_per_month,
    sunrise = as.POSIXct(NA),
    sunset = as.POSIXct(NA)
)

sunrise_sunset_df <- sunrise_sunset_df |> dplyr::mutate(
    sunrise = purrr::map_vec(datetime, ~ suntools::sunriset(
        coords,
        .x,
        crs = sf::st_crs(4326),
        direction='sunrise',
        POSIXct.out=TRUE
    )$time),
    sunset = purrr::map_vec(datetime, ~ suntools::sunriset(
        coords,
        .x,
        crs = sf::st_crs(4326),
        direction='sunset',
        POSIXct.out=TRUE
    )$time),
) 
|> dplyr::mutate(
    sunrise = lubridate::round_date(sunrise, unit = "30minute"),
    sunset = lubridate::round_date(sunset, unit = "30minute")
)

sunrise_sunset_df |>
    print()
```
