---
title: "Generate day/night transition mapping"
---

```{r}
devtools::load_all()
```

We need to evaluate the day/night status for each pixel in the raster. We will fix to UTC so that we can easily compare to the current time on the watch. The steps are:

- Generate a raster to lat/lon coordinate converter
- Generate a grid of times - middle of every month for the next 12 months
  - we presume the yearly variation is tiny
  - the bigger issue is the within-month variability but hey, we are also doing this at an extremely coarse resolution so *shrugs*
- Calculate the sunrise/sunset time for each pixel in the raster
- Convert this into the most minimal representation for the watch to recreate the day/night transition map

We will use `suntools::sunriset()` - takes coords, datetime, crs, direction (sunrise/sunset), and whether to return POSIXct.

```{r}
platform <- "basalt"
debug <- FALSE
centre_lat <- -3.84
centre_lon <- 133.96
```

```{r}
day_image_path <- here::here("blue_pixel","resources",glue::glue("blue_marble_{platform}.png"))
day_image <- magick::image_read(day_image_path)
width <- magick::image_info(day_image)$width
height <- magick::image_info(day_image)$height

# get the available pixels mask (TRUE = non-transparent)
raster_data <- day_image |> as.raster()
# check for transparent pixels via "transparent" string
available_pixels_mask <- raster_data != "transparent"

# for each row, get the min max column index of available pixels
min_max_coords <- apply(available_pixels_mask, 1, function(row) {
    available_indices <- which(row)
    if (length(available_indices) == 0) {
        return(c(NA, NA))
    } else {
        return(c(min(available_indices), max(available_indices)))
    }
}) |> t() |> as.data.frame() |> dplyr::rename(min_x = V1, max_x = V2)

# assert min_max_coords is of length height
stopifnot(nrow(min_max_coords)  == height)

# calculate the radius
radius <- (min_max_coords |> dplyr::mutate(
    dist = (max_x - min_x)
) |> dplyr::select(dist) |> max(na.rm=T)) / 2

```

```{r}
if (debug) {
    # Check on an easily locatable friend Brisbane for Gabbro
    project_to_lat_lon(
        x = 157,
        y = 167,
        lat_ref = centre_lat,
        lon_ref = centre_lon,
        radius = radius,
        width = width,
        height = height
    )
}
```

# Generate the lat/lon coordinates for each pixel in the raster

```{r}
pixel_grid <- expand.grid(
    x = 1:width,
    y = 1:height
) |>
    data.frame() |>
    dplyr::rename(col = x, row = y) |>
    dplyr::mutate(
        is_available = available_pixels_mask[cbind(row, col)],
        coords = purrr::pmap(
            list(col, row), ~ project_to_lat_lon(
                x = ..1,
                y = ..2,
                lat_ref = centre_lat,
                lon_ref = centre_lon,
                radius = radius,
                width = width,
                height = height,
                verbose = FALSE
            )
        ),
        lat = purrr::map_dbl(coords, "lat"),
        lon = purrr::map_dbl(coords, "lon")
    ) |>
    dplyr::select(-coords)

pixel_grid |>
    dplyr::filter(is_available, lat < -27, lon > 153) |>
    head(10)
```


```{r}
if (debug) {
    # show the image as a plot
    library(ggplot2)
    plot_data <- available_pixels_mask |> 
        as.data.frame() |> 
        dplyr::mutate(y = dplyr::row_number()) |>
        tidyr::pivot_longer(-y, names_to = "x", values_to = "value") |> 
        dplyr::mutate(x = as.integer(gsub("V", "", x)))

    ggplot() +
        geom_raster(
            data = plot_data,
            aes(x = x, y = y, fill = value)
        ) +
        scale_fill_manual(values = c("FALSE" = "black", "TRUE" = "blue")) +
        theme_void() +
        theme(legend.position = "none") +
        coord_fixed()
}
```

```{r}
# Get one time per month for the next 12 months
datetime_per_month <- lubridate::ymd_hms(
    "2026-01-01 12:00:00",
    tz = "UTC"
) +
    lubridate::dmonths(0.5) +
    lubridate::dmonths(0:11)

demo_sunset_sunrise <- get_monthly_sunrise_sunset(centre_lat, centre_lon, datetime_per_month)
demo_sunset_sunrise |>
    print()
```

Now we need to get an array int (months, 24*2, row, 2). This will contain for each half hour increment (per month) the farthest left pixel that is in daylight and the farthest right pixel that is in daylight.

```{r}
# we will start by pre-populating this, we know the min and max pixel that is available for each row, so we can start with that and then narrow inwards as we find the sunrise/sunset times

increments_per_day <- 24 * 2
increment_times <- seq(0, 23.5, by = 0.5)

daylight_limits <- array(
    NA_integer_,
    dim = c(
        length(datetime_per_month),
        increments_per_day, # every half hour increment
        height,
        2
    ),
    dimnames = list(
        month = 1:length(datetime_per_month),
        time = increment_times,
        row = 1:height,
        limit = c("left", "right")
    )
)

# populate the limits with the min and max available pixels for each row
daylight_limits[,,,"left"] <- array(
    rep(min_max_coords$max_x |> tidyr::replace_na(width), each = length(datetime_per_month) * increments_per_day),
    dim = c(
        length(datetime_per_month),
        increments_per_day,
        height
    )
)
daylight_limits[,,,"right"] <- array(
    rep(min_max_coords$min_x |> tidyr::replace_na(1), each = length(datetime_per_month) * increments_per_day),
    dim = c(
        length(datetime_per_month),
        increments_per_day,
        height
    )
)
```

```{r}
# now we will iterate over each pixel and calculate the sunrise/sunset times, and then update
avail_pixel_grid <- pixel_grid |>
    dplyr::filter(is_available)
time_dim <- dimnames(daylight_limits)$time |> as.numeric()

pb <- cli::cli_progress_bar(total = nrow(avail_pixel_grid), format = "{cli::pb_bar} {cli::pb_current}/{cli::pb_total} | ETA: {cli::pb_eta}")

for (ii in seq_len(nrow(avail_pixel_grid))) {
    cli::cli_progress_update(id = pb)
    this_row <- avail_pixel_grid$row[ii]
    this_col <- avail_pixel_grid$col[ii]
    ss_times <- get_monthly_sunrise_sunset(
        avail_pixel_grid$lat[ii],
        avail_pixel_grid$lon[ii],
        datetime_per_month
    )

    for (month_idx in ss_times$month) {
        sunrise_time <- ss_times$sunrise[month_idx]
        sunset_time <- ss_times$sunset[month_idx]

        if (is.na(sunrise_time) | is.na(sunset_time)) {
            # if we don't have sunrise/sunset times, then we assume it's either always day or always night, so we skip
            next
        }

        # time mask
        time_mask <- rep(FALSE, length(time_dim))
        if (sunset_time < sunrise_time) {
            # UTC fun 
            time_mask[time_dim <= sunset_time | time_dim >= sunrise_time] <- TRUE
        } else {
            # normal case - day and night
            time_mask[time_dim >= sunrise_time & time_dim <= sunset_time] <- TRUE
        }

        # ugly mutate
        slice_left <- daylight_limits[month_idx, time_mask, this_row, "left"]
        slice_right <- daylight_limits[month_idx, time_mask, this_row, "right"]
        slice_left[slice_left > this_col] <- this_col
        slice_right[slice_right < this_col] <- this_col
        daylight_limits[month_idx, time_mask, this_row, "left"] <- slice_left
        daylight_limits[month_idx, time_mask, this_row, "right"] <- slice_right
    }
}
cli::cli_progress_done(id = pb)

# save to RDS file
saveRDS(daylight_limits, here::here("data", "daylight_limits.rds"))
```


```{r}
daylight_limits <- readRDS(here::here("data", "daylight_limits.rds"))
plot_daylight(
    daylight_limits, month = 4, time = seq(0, 23, by = 2),
    width = width, height = height, radius = radius, tz = "Australia/Sydney"
)
```

# Convert to a binary format that can be read into pebble

From what I understand we need to use `writeBin()` to write a flattened array then if we are strict about the array dims then we will know how to extract the bytes of interest on the watch side. We will use uint16 (2 bytes per int - uint16_t) as uint8 (1 byte per int) as gabbro is 260 width and uint8 maxes out at 255, oof.

I have no idea what this means, it honestly sounds made up, but the endian is "little" for pebble apparently...

```{r}
daylight_limits <- readRDS(here::here("data", "daylight_limits.rds"))

if (all(dim(daylight_limits) == c(12, 48, height, 2))) {
    write_limits_bin(
        daylight_limits,
        platform
    )
} else {
    stop("Unexpected array dimensions: ", paste(dim(daylight_limits), collapse = ", "))
}
```

```{r}
# spot-check a month/hour against the in-memory array
test_month <- 6
test_hour  <- 12
read_back  <- read_limits_bin(
    here::here("blue_pixel", "resources", glue::glue("limits_{platform}.bin")),
    month = test_month, hour = test_hour, height = height
)
time_vals <- seq(0, 23.5, by = 0.5)
time_key  <- dimnames(daylight_limits)$time[which.min(abs(time_vals - test_hour))]
original  <- daylight_limits[test_month, time_key, , ]
if (
    identical(read_back[, "left"],  as.integer(original[, "left"]))
    & identical(read_back[, "right"], as.integer(original[, "right"]))
) {
    message("Verification passed: month=", test_month, ", hour=", test_hour)
} else {
    stop("Verification failed: read back values do not match original array")
}

# print(glue::glue("Original left: {paste(original[, 'left'], collapse = ', ')}"))
# print(glue::glue("Read back left: {paste(read_back[, 'left'], collapse = ', ')}"))
# print(glue::glue("Original right: {paste(original[, 'right'], collapse = ', ')}"))
# print(glue::glue("Read back right: {paste(read_back[, 'right'], collapse = ', ')}"))
```