---
title: "Generate day/night transition mapping"
---

```{r}
devtools::load_all()
```

We need to evaluate the day/night status for each pixel in the raster. We will fix to UTC so that we can easily compare to the current time on the watch. The steps are:

- Generate a raster to lat/lon coordinate converter
- Generate a grid of times - middle of every month for the next 12 months
  - we presume the yearly variation is tiny
  - the bigger issue is the within-month variability but hey, we are also doing this at an extremely coarse resolution from a hand drawn image (I'm no artist) with a basic projection, so *shrugs*
- Calculate the sunrise/sunset time for each pixel in the raster
- Convert this into the most minimal representation for the watch to recreate the day/night transition map

We will use `suntools::sunriset()` - takes coords, datetime, crs, direction (sunrise/sunset), and whether to return POSIXct.

```{r}
platform <- "basalt"
debug <- FALSE
# Crudely I have estimated the centre pixel of the image
centre_lat <- -3.84
centre_lon <- 133.96
```

## Load the image

```{r}
day_image_path <- here::here("blue_pixel","resources",glue::glue("blue_marble_{platform}.png"))
day_image <- magick::image_read(day_image_path)
width <- magick::image_info(day_image)$width
height <- magick::image_info(day_image)$height

# get the available pixels mask (TRUE = non-transparent)
raster_data <- day_image |> as.raster()
# check for transparent pixels via "transparent" string
available_pixels_mask <- raster_data != "transparent"

# for each row, get the min max column index of available pixels
min_max_coords <- apply(available_pixels_mask, 1, function(row) {
    available_indices <- which(row)
    if (length(available_indices) == 0) {
        return(c(NA, NA))
    } else {
        return(c(min(available_indices), max(available_indices)))
    }
}) |> t() |> as.data.frame() |> dplyr::rename(min_x = V1, max_x = V2)

# assert min_max_coords is of length height
stopifnot(nrow(min_max_coords)  == height)

# calculate the radius
radius <- (min_max_coords |> dplyr::mutate(
    dist = (max_x - min_x)
) |> dplyr::select(dist) |> max(na.rm=T)) / 2

```

```{r}
if (debug) {
    # Check on an easily locatable friend Brisbane for Gabbro
    project_to_lat_lon(
        x = 157,
        y = 167,
        lat_ref = centre_lat,
        lon_ref = centre_lon,
        radius = radius,
        width = width,
        height = height
    )
}
```

# Generate the lat/lon coordinates for each pixel in the raster

```{r}
pixel_grid <- expand.grid(
    x = 1:width,
    y = 1:height
) |>
    data.frame() |>
    dplyr::rename(col = x, row = y) |>
    dplyr::mutate(
        is_available = available_pixels_mask[cbind(row, col)],
        coords = purrr::pmap(
            list(col, row), ~ project_to_lat_lon(
                x = ..1,
                y = ..2,
                lat_ref = centre_lat,
                lon_ref = centre_lon,
                radius = radius,
                width = width,
                height = height,
                verbose = FALSE
            )
        ),
        lat = purrr::map_dbl(coords, "lat"),
        lon = purrr::map_dbl(coords, "lon")
    ) |>
    dplyr::select(-coords)

pixel_grid |>
    dplyr::filter(is_available, lat < -27, lon > 153) |>
    head(10)
```


```{r}
if (debug) {
    # show the image as a plot
    library(ggplot2)
    plot_data <- available_pixels_mask |> 
        as.data.frame() |> 
        dplyr::mutate(y = dplyr::row_number()) |>
        tidyr::pivot_longer(-y, names_to = "x", values_to = "value") |> 
        dplyr::mutate(x = as.integer(gsub("V", "", x)))

    ggplot() +
        geom_raster(
            data = plot_data,
            aes(x = x, y = y, fill = value)
        ) +
        scale_fill_manual(values = c("FALSE" = "black", "TRUE" = "blue")) +
        theme_void() +
        theme(legend.position = "none") +
        coord_fixed()
}
```

## Check the sunrise/sunset time function

```{r}
# Get one time per month for the next 12 months
datetime_per_month <- lubridate::ymd_hms(
    "2026-01-01 12:00:00",
    tz = "UTC"
) +
    lubridate::dmonths(0.5) +
    lubridate::dmonths(0:11)

demo_sunset_sunrise <- get_monthly_sunrise_sunset(centre_lat, centre_lon, datetime_per_month)
demo_sunset_sunrise |>
    print()
```

Create a full boolean pixel array `(months, time_slots, rows, cols)` where TRUE = daylight. This gives us a complete unconstrained representation we can visualise and smooth before compressing to left/right limits later.

```{r}
increments_per_day <- 24 * 2
increment_times <- seq(0, 23.5, by = 0.5)

daylight_bool <- array(
    NA,
    dim = c(
        length(datetime_per_month),
        increments_per_day,
        height,
        width
    ),
    dimnames = list(
        month = 1:length(datetime_per_month),
        time = increment_times,
        row = 1:height,
        col = 1:width
    )
)
message("daylight_bool size: ", format(object.size(daylight_bool), units = "MB"))
```

```{r}
avail_pixel_grid <- pixel_grid |>
    dplyr::filter(is_available) |>
    dplyr::filter(!is.na(lat) & !is.na(lon))

message("Processing ", nrow(avail_pixel_grid), " valid pixels (with non-NA lat/lon)")

time_dim <- dimnames(daylight_bool)$time |> as.numeric()

pb <- cli::cli_progress_bar(total = nrow(avail_pixel_grid), format = "{cli::pb_bar} {cli::pb_current}/{cli::pb_total} | ETA: {cli::pb_eta}")

for (ii in seq_len(nrow(avail_pixel_grid))) {
    cli::cli_progress_update(id = pb)
    this_row <- avail_pixel_grid$row[ii]
    this_col <- avail_pixel_grid$col[ii]

    # Mark this pixel as processed: default to FALSE (night) for all slots.
    # Unprocessed pixels (NA lat/lon, outside circle) remain NA.
    daylight_bool[, , this_row, this_col] <- FALSE

    ss_times <- get_monthly_sunrise_sunset(
        avail_pixel_grid$lat[ii],
        avail_pixel_grid$lon[ii],
        datetime_per_month
    )

    for (month_idx in ss_times$month) {
        sunrise_time <- ss_times$sunrise[month_idx]
        sunset_time <- ss_times$sunset[month_idx]

        if (is.na(sunrise_time) | is.na(sunset_time)) {
            next  # perpetual night - already FALSE from above
        }

        # time mask
        time_mask <- rep(FALSE, length(time_dim))
        if (sunset_time < sunrise_time) {
            time_mask[time_dim <= sunset_time | time_dim >= sunrise_time] <- TRUE
        } else {
            time_mask[time_dim >= sunrise_time & time_dim <= sunset_time] <- TRUE
        }

        daylight_bool[month_idx, time_mask, this_row, this_col] <- TRUE
    }
}
cli::cli_progress_done(id = pb)

message("Main loop done. Checking for NA foreground pixels to fill...")
```

Sometimes there are pixels that can't be dealt with (weird projection or suntools will not provide an output for them). So we'll just adopt an approach of filling any NA foreground pixels with the nearest neighbour that has a valid day/night pattern.

```{r}
#save the intermediate result before filling NA pixels, just for posterity
saveRDS(daylight_bool, here::here("data", glue::glue("daylight_bool_pre_fill_{platform}.rds")))
message("Saved intermediate daylight_bool (pre-fill) to data/daylight_bool_pre_fill_", platform, ".rds")

# Nearest-neighbor fill for foreground pixels that have NA (bad projection edges)
# These are pixels marked available in the image but project_to_lat_lon returned NA
na_fg_pixels <- pixel_grid |>
    dplyr::filter(
        is_available,
        purrr::map2_lgl(row, col, ~ anyNA(daylight_bool[, , .x, .y]))
    )

if (nrow(na_fg_pixels) > 0) {
    message("Filling ", nrow(na_fg_pixels), " NA foreground pixels via nearest-neighbor")
    valid_coords <- avail_pixel_grid[, c("row", "col")]

    for (i in seq_len(nrow(na_fg_pixels))) {
        target_row <- na_fg_pixels$row[i]
        target_col <- na_fg_pixels$col[i]

        # Find nearest valid pixel by Euclidean distance
        dists <- (valid_coords$row - target_row)^2 + (valid_coords$col - target_col)^2
        nearest_idx <- which.min(dists)
        nearest_row <- valid_coords$row[nearest_idx]
        nearest_col <- valid_coords$col[nearest_idx]

        # Copy all (month, time) values from nearest valid pixel
        daylight_bool[, , target_row, target_col] <- daylight_bool[, , nearest_row, nearest_col]
    }
    message("Nearest-neighbor fill complete")
} else {
    message("No NA foreground pixels to fill")
}

saveRDS(daylight_bool, here::here("data", glue::glue("daylight_bool_{platform}.rds")))
message("Saved daylight_bool to data/daylight_bool_", platform, ".rds")
```

## Checking plots

```{r}
daylight_bool <- readRDS(here::here("data", glue::glue("daylight_bool_{platform}.rds")))
plot_daylight_bool(
    daylight_bool, month = 6, times = seq(0, 23, by = 2),
    width = width, height = height, radius = radius, tz = "Australia/Sydney"
)
# Quick summary of a single time slice
slice <- daylight_bool[6, 9, , 1:50]
message("Month 6, time 14.0: ", sum(slice), " lit pixels out of ", length(slice))
```

# Convert to a binary format that can be read into pebble

From what I understand we need to use `writeBin()` to write a flattened array then if we are strict about the array dims then we will know how to extract the bytes of interest on the watch side. We will use uint16 (2 bytes per int - uint16_t) as uint8 (1 byte per int) as gabbro is 260 width and uint8 maxes out at 255, oof.

I have no idea what this means, it honestly sounds made up, but the endian is "little" for pebble apparently...

```{r}
daylight_bool <- readRDS(here::here("data", glue::glue("daylight_bool_{platform}.rds")))
daylight_limits <- compress_bool_to_limits(daylight_bool, min_max_coords)
message("Compressed to limits array: ", paste(dim(daylight_limits), collapse = " x "))
saveRDS(daylight_limits, here::here("data", glue::glue("daylight_limits_{platform}.rds")))
```

```{r}
# Verify the compression/decompression visually
plot_daylight(
    daylight_limits, month = 6, times = seq(0, 23, by = 2),
    width = width, height = height, radius = radius,
    min_max_coords = min_max_coords, tz = "Australia/Sydney"
)
```

```{r}
# Detailed single-time plot showing boundary pixels
plot_daylight_single(
    daylight_limits, month = 6, time = 16,
    width = width, height = height, radius = radius,
    min_max_coords = min_max_coords, tz = "Australia/Sydney"
)

# Quick summary of a single time slice
slice <- daylight_limits[6, 13, ,]
```

## Write to binary for peb

```{r}
stored_rows <- ceiling(height / 2)
if (all(dim(daylight_limits) == c(12, 48, stored_rows, 3))) {
    write_limits_bin(
        daylight_limits,
        platform
    )
    message("Binary written. Expected size: ", 12 * 48 * stored_rows * 3, " bytes")
} else {
    stop("Unexpected array dimensions: ", paste(dim(daylight_limits), collapse = ", "))
}
```

Briefly check all went well

```{r}
# spot-check a month/hour against the in-memory array
test_month <- 6
test_hour  <- 12
read_back  <- read_limits_bin(
    here::here("blue_pixel", "resources", glue::glue("limits_{platform}.bin")),
    month = test_month, hour = test_hour, height = stored_rows
)
time_vals <- seq(0, 23.5, by = 0.5)
time_key  <- dimnames(daylight_limits)$time[which.min(abs(time_vals - test_hour))]
original  <- daylight_limits[test_month, time_key, , ]
if (
    identical(read_back[, "left"],  as.integer(original[, "left"]))
    & identical(read_back[, "right"], as.integer(original[, "right"]))
    & identical(read_back[, "left2"], as.integer(original[, "left2"]))
) {
    message("Verification passed: month=", test_month, ", hour=", test_hour)
} else {
    stop("Verification failed: read back values do not match original array")
}
```

