---
title: "Generate day/night transition mapping"
---

```{r}
devtools::load_all()
```

We need to evaluate the day/night status for each pixel in the raster. We will fix to UTC so that we can easily compare to the current time on the watch. The steps are:

- Generate a raster to lat/lon coordinate converter
- Generate a grid of times - middle of every month for the next 12 months
  - we presume the yearly variation is tiny
  - the bigger issue is the within-month variability but hey, we are also doing this at an extremely coarse resolution so *shrugs*
- Calculate the sunrise/sunset time for each pixel in the raster
- Convert this into the most minimal representation for the watch to recreate the day/night transition map

We will use `suntools::sunriset()` - takes coords, datetime, crs, direction (sunrise/sunset), and whether to return POSIXct.

```{r}
platform <- "basalt"
debug <- FALSE
centre_lat <- -3.84
centre_lon <- 133.96
```

```{r}
day_image_path <- here::here("blue_pixel","resources",glue::glue("blue_marble_{platform}.png"))
day_image <- magick::image_read(day_image_path)
width <- magick::image_info(day_image)$width
height <- magick::image_info(day_image)$height

# get the available pixels mask (TRUE = non-transparent)
raster_data <- day_image |> as.raster()
# check for transparent pixels via "transparent" string
available_pixels_mask <- raster_data != "transparent"

# for each row, get the min max column index of available pixels
min_max_coords <- apply(available_pixels_mask, 1, function(row) {
    available_indices <- which(row)
    if (length(available_indices) == 0) {
        return(c(NA, NA))
    } else {
        return(c(min(available_indices), max(available_indices)))
    }
}) |> t() |> as.data.frame() |> dplyr::rename(min_x = V1, max_x = V2)

# assert min_max_coords is of length height
stopifnot(nrow(min_max_coords)  == height)

# calculate the radius
radius <- (min_max_coords |> dplyr::mutate(
    dist = (max_x - min_x)
) |> dplyr::select(dist) |> max(na.rm=T)) / 2

```

```{r}
if (debug) {
    # Check on an easily locatable friend Brisbane for Gabbro
    project_to_lat_lon(
        x = 157,
        y = 167,
        lat_ref = centre_lat,
        lon_ref = centre_lon,
        radius = radius,
        width = width,
        height = height
    )
}
```

# Generate the lat/lon coordinates for each pixel in the raster

```{r}
pixel_grid <- expand.grid(
    x = 1:width,
    y = 1:height
) |>
    data.frame() |>
    dplyr::rename(col = x, row = y) |>
    dplyr::mutate(
        is_available = available_pixels_mask[cbind(row, col)],
        coords = purrr::pmap(
            list(col, row), ~ project_to_lat_lon(
                x = ..1,
                y = ..2,
                lat_ref = centre_lat,
                lon_ref = centre_lon,
                radius = radius,
                width = width,
                height = height,
                verbose = FALSE
            )
        ),
        lat = purrr::map_dbl(coords, "lat"),
        lon = purrr::map_dbl(coords, "lon")
    ) |>
    dplyr::select(-coords)

pixel_grid |>
    dplyr::filter(is_available, lat < -27, lon > 153) |>
    head(10)
```


```{r}
if (debug) {
    # show the image as a plot
    library(ggplot2)
    plot_data <- available_pixels_mask |> 
        as.data.frame() |> 
        dplyr::mutate(y = dplyr::row_number()) |>
        tidyr::pivot_longer(-y, names_to = "x", values_to = "value") |> 
        dplyr::mutate(x = as.integer(gsub("V", "", x)))

    ggplot() +
        geom_raster(
            data = plot_data,
            aes(x = x, y = y, fill = value)
        ) +
        scale_fill_manual(values = c("FALSE" = "black", "TRUE" = "blue")) +
        theme_void() +
        theme(legend.position = "none") +
        coord_fixed()
}
```

```{r}
# Get one time per month for the next 12 months
datetime_per_month <- lubridate::ymd_hms(
    "2026-01-01 12:00:00",
    tz = "UTC"
) +
    lubridate::dmonths(0.5) +
    lubridate::dmonths(0:11)

demo_sunset_sunrise <- get_monthly_sunrise_sunset(centre_lat, centre_lon, datetime_per_month)
demo_sunset_sunrise |>
    print()
```

Now we need to get an array int (months, 24*2, row, 2). This will contain for each half hour increment (per month) the farthest left pixel that is in daylight and the farthest right pixel that is in daylight.

```{r}
# we will start by pre-populating this, we know the min and max pixel that is available for each row, so we can start with that and then narrow inwards as we find the sunrise/sunset times

increments_per_day <- 24 * 2
increment_times <- seq(0, 23.5, by = 0.5)

daylight_limits <- array(
    NA_integer_,
    dim = c(
        length(datetime_per_month),
        increments_per_day, # every half hour increment
        height,
        2
    ),
    dimnames = list(
        month = 1:length(datetime_per_month),
        time = increment_times,
        row = 1:height,
        limit = c("left", "right")
    )
)

# populate the limits with the min and max available pixels for each row
daylight_limits[,,,"left"] <- array(
    rep(min_max_coords$max_x |> tidyr::replace_na(width), each = length(datetime_per_month) * increments_per_day),
    dim = c(
        length(datetime_per_month),
        increments_per_day,
        height
    )
)
daylight_limits[,,,"right"] <- array(
    rep(min_max_coords$min_x |> tidyr::replace_na(1), each = length(datetime_per_month) * increments_per_day),
    dim = c(
        length(datetime_per_month),
        increments_per_day,
        height
    )
)
```

```{r}
# now we will iterate over each pixel and calculate the sunrise/sunset times, and then update
avail_pixel_grid <- pixel_grid |>
    dplyr::filter(is_available)
time_dim <- dimnames(daylight_limits)$time |> as.numeric()

pb <- cli::cli_progress_bar(total = nrow(avail_pixel_grid), format = "{cli::pb_bar} {cli::pb_current}/{cli::pb_total} | ETA: {cli::pb_eta}")

for (ii in seq_len(nrow(avail_pixel_grid))) {
    cli::cli_progress_update(id = pb)
    this_row <- avail_pixel_grid$row[ii]
    this_col <- avail_pixel_grid$col[ii]
    ss_times <- get_monthly_sunrise_sunset(
        avail_pixel_grid$lat[ii],
        avail_pixel_grid$lon[ii],
        datetime_per_month
    )

    for (month_idx in ss_times$month) {
        sunrise_time <- ss_times$sunrise[month_idx]
        sunset_time <- ss_times$sunset[month_idx]

        if (is.na(sunrise_time) | is.na(sunset_time)) {
            # if we don't have sunrise/sunset times, then we assume it's either always day or always night, so we skip
            next
        }

        # time mask
        time_mask <- rep(FALSE, length(time_dim))
        if (sunset_time < sunrise_time) {
            # UTC fun 
            time_mask[time_dim >= sunset_time | time_dim <= sunrise_time] <- TRUE
        } else {
            # normal case - day and night
            time_mask[time_dim >= sunrise_time & time_dim <= sunset_time] <- TRUE
        }

        # ugly mutate
        slice_left <- daylight_limits[month_idx, time_mask, this_row, "left"]
        slice_right <- daylight_limits[month_idx, time_mask, this_row, "right"]
        slice_left[slice_left > this_col] <- this_col
        slice_right[slice_right < this_col] <- this_col
        daylight_limits[month_idx, time_mask, this_row, "left"] <- slice_left
        daylight_limits[month_idx, time_mask, this_row, "right"] <- slice_right
    }
}
cli::cli_progress_done(id = pb)

# save to RDS file
saveRDS(daylight_limits, here::here("data", "daylight_limits.rds"))
```


```{r}
daylight_limits <- readRDS(here::here("data", "daylight_limits.rds"))
plot_daylight(
    daylight_limits, month = 6, time = seq(0, 23, by = 2),
    width = width, height = height, radius = radius
)
```